Ornaments as indicators of Social change before and after European contact at Kiwulan, Northeastern Taiwan

### 2019-07-05 analysis/paper/paper.Rmd
added map and references, and revised introduction

1. background of the topic

Taiwan island became part of a global trading network during the early 17th century because of western European expansion that connected European and Asia. Compared to other places in Asia and Oceania, the impact of European colonialism on indigenous communities in East Asia appears to have been much less pronounced. Direct colonial rule was rare and limited, but the question of long-lasting indirect impacts on local indigenous communities remains unanswered. Indirect effects of colonialism is emphasized increasingly to discuss the colonial impact on the indigenous people in the periphery of colonial control [@TorrenceandClarke2000; @Trabert2017]. In many parts of the world, foreign trade goods introduced into local indigenous societies in a colonial or imperial contexts might cause significant transformations of indigenous economic, cultural, and socio-political systems [@Dietler1997; @Junker1993; @Mitchell2000; @Silliman2005]. Contrary to as passive receptors of imperial power, the agency of indigenous people in the colonial contexts has been broadly discussed to emphasize the ability of active adoptions of foreign effects, negotiation between colonized and colonizer, and resistance of indigenous people through daily cultural practices [@Dietler2015; @Given2004; @Rubertone2000; @Silliman2001; @Torrence2000; @TorrenceandClarke2000; @Voss2005]. Studying the consumption of foreign goods is one way to understand indigenous experience of the colonial encounter and explore the agency of indigenous people by examining how those goods were distributed and used when obtained [@Dietler2005; @Mullins2011; @Scaramelli2005].

Taiwan was colonized by the Spanish and the Dutch subsequently from 1624 to 1662, where northeastern Taiwan, Yilan, is an especially unique example for exploring indirect effects due to its physically isolated location that made it not easily accessible to the Europeans and the colonial control was less compared to other parts of this region [@Andrade2007; @Kang2012]. However, the sea provides a way through which indigenous people could interact with other groups from different regions, including other native indigenous groups, foreign traders such as the Han people of China, and Europeans. Large-scale and frequent trade activities brought a wide variety of trade objects into indigenous communities in Yilan through local regional exchange network. Among those trade goods, ornament such as glass and stone beads was one of the common foreign goods found at the settlement sites during this period [@Chen2007; @Li2014; @NMTH2005; @Wang2007]. Body adornment is usually viewed as a signal in indigenous social life to represent and reinforce their status through the body as the scene to display [@Joyce2005; @Scaramelli2005]. The consumption of trade beads in Southeast Asia during the Iron Age were usually associated with increasing social stratification or socio-political complexity that influenced [@Bellina2014; @Carter2016; @Francis2002; @Theunissen2000].

identify the value of the beads

@Theunissen2000 <---This has never been explored in any depth despite the fact it has been recognized for some time that the introduction of new prestige goods from external sources may cause restructuring of existing prestige goods exchange and subsequent increases in social complexity (Higham 1984: 250-4). --->

@Mullins2011 <---suggest that beads were one class of goods that continued and embellished indigenous value systems that stressed the powers embedded in shiny objects such as quartz crystals, and the European use of beads in items such as rosaries may have reinforced their powers in the eyes of indigenous people.--->

@Carter2016 <---has also proposed that these lower quality beads may instead be related to intensified trade with South Asia and the importation of mass-produced beads from this location. --->

My research will explore the evidence at Kiwulan by addressing the question of whether European contact and influence on local trade networks stimulated an increase in social inequality among the inhabitants of
Kiwulan. In this project, I intend to test the hypothesis that European contact in northeastern Taiwan stimulated a change in indigenous social organization at Kiwulan.

2. debate or problem have not solved yet
3. how this case can answer it
4. what are you going to do

We ask whether colonial influence on indigenous populations can be detected in ornaments recovered from the archaeological record at Kiwulan (1400-1900 AD), a large Iron Age settlement in northeastern Taiwan[@Chen2007]. Our hypothesis is that there was a greater diversity of ornaments types and materials at Kiwulan after European contact due to the large scale exchange network. We hypothesize that after Chinese contact there was a decline in the production, use and discard of ornaments at Kiwulan. We hypothesize that there are clustered distribution of ornaments at Kiwulan relating to these international contacts.

@@

Foreign ornaments including glass beads, stone beads, and metal objects first appeared in Taiwan around 1,800 years ago through the regional exchange network in Southeast Asia [@Wang2007]. Those new types of ornaments gradually replaced the earlier jade ornamental system in Taiwan during the transition from the Neolithic to the Iron Age. We observed a greater quantity and variety of foreign goods in the early 17th century during the European colonization period that involved the whole island in an international trade network on a large scale.

Prior to the contact with the Europeans, there were small-scale regional exchange networks between China and Taiwan since in the Neolithic period. Also, northern Taiwan had been involved in long-distance networks of East Asia since 14th century, and encountered a large wave of Han Chinese migration in the 19th century. Northeastern Taiwan can provide a good example to explore the indirect impacts on indigenous societies caused by different foreign groups in different periods by taking a long-term perspective research.

@@

Kiwulan (Figure 1) is located at Yilan, north of Lanyang Plain in northeastern Taiwan, surrounded by the Central Mountain Range in the west and the Pacific Ocean in the east. Kiwulan was excavated from 2001 to 2004, and the total excavation area was 3,814 m2. Most beads were found during the excavation and some of them through screens with 20 mm and 1.5 mm mesh. The archaeological evidence includes a rich amount of artifacts, burials, middens, post holes, wooden pillars, and stone structures. The chronology of Kiwulan can be divided into Lower Layer Culture (700 - 1200 AD) and Upper Layer Culture (1400 - 1900 AD) with a sterile layer in between based on a series of 32 radiocarbon dates[a], which coordinate to the late Iron Age and Proto-historical period in Taiwan. This paper focuses on the Upper Layer Culture which overlaps with the contact periods with the Europeans and Chinese. It provides a good examples to discuss the foreign impacts on local indigenous society through body ornaments that usually convey social and cultural meaning by displaying.

>>>>

Taiwan had encountered multiple episodes of culture contact with China and Southeast Asia since Neolithic and became part of a global trading network during the early 17th century because of western European expansion that connected European and Asia. Compared to other places in Asia and Oceania, the impact of European colonialism on indigenous communities in East Asia appears to have been much less pronounced. Direct colonial rule was rare and limited, but the question of long-lasting indirect impacts on local indigenous communities remains unanswered. Indirect effects of colonialism is emphasized increasingly to discuss the colonial impact on the indigenous people in the periphery of colonial control [@TorrenceandClarke2000; @Trabert2017]. In many parts of the world, foreign trade goods introduced into local indigenous societies in a colonial or imperial contexts might cause significant transformations of indigenous economic, cultural, and socio-political systems [@Dietler1997; @Junker1993; @Mitchell2000; @Silliman2005]. Contrary to as passive receptors of imperial power, the agency of indigenous people in the colonial contexts has been broadly discussed to emphasize the ability of active adoptions of foreign effects, negotiation between colonized and colonizer, and resistance of indigenous people through daily cultural practices [@Dietler2015; @Given2004; @Rubertone2000; @Silliman2001; @Torrence2000; @TorrenceandClarke2000; @Voss2005]. Studying the consumption of foreign goods is one way to understand indigenous experience of the colonial encounter and explore the agency of indigenous people by examining how those goods were distributed and used when obtained [@Dietler2005; @Mullins2011; @Scaramelli2005].

Despite colonization of Taiwan by the Spanish and the Dutch from 1624 to 1662, little has been discovered in archaeology about the The reaction of indigenous people to the European colonization. Northeastern Taiwan, Yilan, is an especially unique example for exploring indirect effects due to its physically isolated location that made it not easily accessible to the Europeans and the colonial control was less compared to other parts of this region [@Andrade2007; @Kang2012]. Although the location is relatively peripheral, the sea provides a way through which indigenous people could interact with other groups from different regions, including other native indigenous groups, foreign traders such as the Han people of China, and Europeans. Large-scale and frequent trade activities brought a wide variety of trade objects into indigenous communities in Yilan through local regional exchange network. Among those trade goods, ornament such as glass and stone beads was one of the common foreign goods found at the settlement sites in this region [@Chen2007; @Li2014; @NMTH2005]. Body adornment is usually viewed as a signal in indigenous social life to represent and reinforce their status through the body as the scene to display [@Joyce2005; @Scaramelli2005]. For example, the consumption of stone beads in Southeast Asia during Iron Age were usually associated with increasing social stratification or socio-political complexity [@Bellina2014; @Carter2016; @Francis2002; @Theunissen2000].

In this paper, I explore the evidence from a site spans from late Iron Age to historical period by addressing the question of whether colonial influence on indigenous populations can be detected in ornaments recovered from the archaeological record at Kiwulan (1400-1900 AD), a large Iron Age settlement in northeastern Taiwan [@Chen2007]. Prior to the contact with the Europeans in the early 17th century, northern Taiwan had been involved in long-distance networks of East Asia since 14th century, and later encountered a large wave of Han Chinese migration in the 19th century [@Chen2005]. Northeastern Taiwan provides a good example to explore the indirect impacts on indigenous societies caused by comparing different episodes of culture contact based on a long-term perspective. My hypotheses are that there was a greater diversity of ornaments types and materials at Kiwulan after European contact due to the large scale exchange network, and the consumption of ornaments present some spatial patterns that indicate increasing social inequality. I also hypothesize that the use of ornaments declined during Chinese contact period.

@@

Foreign ornaments including glass beads, stone beads, and metal objects first appeared in Taiwan around 1,800 BCE through the regional exchange network in Southeast Asia [@Wang2007]. Those new types of ornaments gradually replaced the earlier jade ornamental system in Taiwan during the transition from the Neolithic to the Iron Age. We observed a greater quantity and variety of foreign goods in the early 17th century during the European colonization period that involved the whole island in an international trade network on a large scale.

@@

```{r, KWL-map, fig.cap = "Kiwulan location in Southeast Asia", out.width= "400px", echo=FALSE}
knitr::include_graphics("kiwulan-location-map.jpeg")
```

Kiwulan \@ref(fig:KWL-map) is located at Yilan, north of Lanyang Plain in northeastern Taiwan, surrounded by the Central Mountain Range in the west and the Pacific Ocean in the east. Kiwulan was excavated from 2001 to 2004, and the total excavation area was 3,814 m2. Most beads were found during the excavation and some of them through screens with 20 mm and 1.5 mm mesh. The archaeological evidence includes a rich amount of artifacts, burials, middens, post holes, wooden pillars, and stone structures. The chronology of Kiwulan can be divided into Lower Layer Culture (700 - 1200 AD) and Upper Layer Culture (1400 - 1900 AD) with a sterile layer in between based on a series of 32 radiocarbon dates[a], which coordinate to the late Iron Age and Proto-historical period in Taiwan. This paper focuses on the Upper Layer Culture which overlaps with the contact periods with the Europeans and Chinese. It provides a good examples to discuss the foreign impacts on local indigenous society through body ornaments that usually convey social and cultural meaning by displaying.
